Scalar Feedback based Joint Time-Frequency Precoder interpolation for MIMO-OFDM Systems

### 2020-03-03 paper/response_letter.tex

\newpage

\textbf{Editor comments}\\

\textbf{There are several mistakes and typos in the manuscript, as
  also pointed out by both reviewers. Moreover, it is mentioned that
  the temporal correlation based and frequency correlation based
  precoder tracking have been discussed in [8] and [9]. Thus the
  differences between the proposed method and the methods in [8] and
  [9] should be explained and compared, so as to make the
  contributions more clear.}

\textbf{Response:} Thank you for pointing this out. The temporal
correlation based precoder tracking discussed in [8] and [9] have
considered about precoder tracking in time and frequency respectively,
albeit separately, without any joint interpolation.  We have
approached this differently by performing prediction in the time
domain and using this for joint interpolation on the time-freqeuncy
grid. This helps build a framework for reducing the total bit
bandwidth by a significant amount while maintaining similar performance
in terms of BER.  We have altered the text in the introduction to make
this point more clear,  as follows:

@@

Thank you for this comment. Although we simulate over 10 realizations,
the channel evolves beginning with each realization, resulting in
sufficient averaging over various channel characteristics. Therefore,
the results that we present represent simulation over a sufficiently
large number of bits transmitted over enough channel variations. We
have now altered the text in the simulation section to reflect this as
follows:

\emph{The BER achieved by the proposed predictive quantization based
  precoder is compared against the ideal precoder for 100 channel
  evolutions in each run and averaged over 10 channel realizations
  (translating to averaging over $125,000$ bits for 64-point DFT and
  $2\times 10^6$ bits for the 1024-point DFT case). The error
  properties obtained by simulating transmission over these varying
  channels is sufficient to obtain good average performance
  properties.}

\textbf{5. In the manuscript, the time-frequency correlation is used for precoder tracking. Because it is mentioned in Section I that the precoder tracking have been discussed in [8] (based on temporal correlation) and [9] (based on frequency correlation), the difference among the proposed method and the methods in [8] and [9] should be explained and compared in the manuscript.}

In addition, the temporal correlation based precoder tracking
discussed in [8] and [9] have considered about precoder tracking in
time and frequency respectively, albeit separately, without any joint
interpolation.  We have approached this differently by performing
prediction in the time domain and using this for joint interpolation
on the time-freqeuncy grid. This helps build a framework for reducing
the total bit bandwidth by a significant amount while maintaining
similar performance in terms of BER.  We have altered the text in the
introduction to make this point more clear, as follows:

This is in addition to various changes in other sections of the paper
that further emphasize the advantages of our approach.

### 2020-03-04 paper/main.tex
Add changes

%package list
\IEEEoverridecommandlockouts
\let\labelindent\relax
% \usepackage{breqn}
\usepackage{enumitem}
\usepackage{fleqn}
\usepackage{cite}
\usepackage{graphicx}
\usepackage[varg]{newtxmath}
\graphicspath{ {images/} }
\usepackage{pdfpages}
\usepackage{wrapfig}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{lettrine}
\usepackage{amsmath}
% \usepackage{titlesec}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{float}
\usepackage[font={footnotesize}]{caption}
\usepackage[numbers,sort,square,compress]{natbib}
\usepackage[para]{footmisc}
\usepackage{xcolor}
\setlength{\belowcaptionskip}{-12pt}
\newcommand{\highlight}[1]{%
  \colorbox{red!50}{$\displaystyle#1$}}

@@

The parameterization approach described above is related to the channel
state information feedback present in several wireless standards,
including 802.11ac~\cite{lou2013comparison}, although adaptive
quantization and feedback are not discussed. Further
sections show that the use of adaptive approaches can greatly enhance
performance with only very minor modifications to the system.
\subsection{Differential quantization and channel tracking}
\label{quantiz}
For channels that vary slowly with time, adaptive differential
quantization is an effective method for tracking parameters over time
using very few bits (often just one bit per parameter). For example,
while tracking a slowly varying discrete random process $x_n$ at time
$n$, the one bit can be used to determine the direction in which to
move in order to track the parameter. This can be denoted by
$\beta_{n} = \mbox{sign}(x_{n} - \hat{x}_{n-1})$, where $x_n$ is the
unquantized (accurate) value and $\hat{x}_{n-1}$ is the quantized
value. Further, the step size for the move can be adapted based on the
speed of channel variation as well the accuracy needed. If $\Delta_n$
is the step size, we define our adaptive quantizer as
\begin{equation}
\hat{x}_{n} = \hat{x}_{n-1} + \beta_{n}\Delta_{n} \mbox{ where }
\label{delta_eqn}
\Delta_{n} = \begin{cases}
    M \Delta_{n-1}, & \text{if $\beta_{n} = \beta_{n-1}$}\\
    \Delta_{n-1}/M , & \text{if $\beta_{n} \neq \beta_{n-1}$}.
  \end{cases}
\end{equation}
where $M$ is the scaling factor used during the positive and negative
moves. We initialize $\Delta_1$ as $\Delta_1 = |x_{2}-\hat{x}_1|$. We
note that this is performed separately for each of the quantized
parameters.

@@

\section{Simulation and discussion}
\label{section3}
We now present simulations of the scalar feedback based joint
time-frequency predictive quantization outlined in the previous
section. In particular, we analyze the performance of the proposed
quantization and interpolation method for time-varying MIMO-OFDM
systems. The channel is simulated using the typical Jakes Model
for the COST207 channel power delay
profiles~\cite{molisch2006cost259,cost1989cost}. The BER achieved by
the proposed predictive quantization based precoder is compared
against the ideal precoder for 100 channel evolutions in each run and
averaged over 10 channel realizations (translating to averaging over $250,000$ bits). The error properties obtained by simulating transmission over these varying channels is sufficient to obtain good average performance properties. Simulations are performed for
normalised Doppler values of $f_dT_s = 3.41\times 10^{-6}$ (corresponding
to a velocity of 37.2 km/h for $f_c = 1$ GHz) with 1024 subcarriers, and with
$f_dT_s = 1.56 \times 10^{-5}$ (velocity of 172.8 km/h) with 64
subcarriers. The sampling time period $T_s$ was $5\times10^{-8}$
s. For both the situations, we considered $N_T=4$ transmit antennas
and and $N_R=2$ receive antennas. Therefore, using the approach
outlined in~\cite{4114278}, the parameters that determine the precoder is
$N_{R}(2N_{T} - N_R) = 12$, with five $\theta_{k,l}$s and and seven
$\phi_{k,l}$s. We assume that the spacing between fed back subcarrier
indices $p$ to be $33$ and $9$ for the slow and fast channel profiles
respectively (i.e. for the slower channels with 1024 subcarriers, the
subcarriers indices fed back would be $0, 33, 66, \ldots 1023$ and for
the faster channel, with 64 subcarriers, $0, 9, 18, \ldots 63$ would
be fed back).

For the first and the second simulation time instances, since the
precoder is not available at the transmitter, we use 2 bits for
initializing each parameter for effective representation of the
precoder. Therefore, the number of bits for each subcarrier =
$10\times 2$ = 20 (we note that $\phi_0 = \phi_4 = 0$ due to the
non-uniqueness of the SVD). Quantization is performed as given in
Section~\ref{section2}. We note that this is not significant, since,
amortized over long durations, this number becomes small. Subsequent
quantization takes place in the time-domain using one bit for each
parameter. Since the bit budget is small, if the channel varies
sufficiently slowly, the $\theta_{k,l}$ and the $\phi_{k,l}$ parameters
are fed back alternately, while retaining the unsent parameters as
their previous values, as shown in
Fig.~\ref{fig:adpm-fig}.
% Here, $\bf{\Theta}$ is the collection of all
% $\theta_l$s and $\bf{\Phi}$ is the collection of all $\phi_i$s.
This brings down the bit budget to an average of 5 bits for
$\bf{\Theta}$ and 5 bits for $\bf{\Phi}$ over alternate feedback
instances, thereby resulting in an average feedback budget of 5 bits,
as opposed to the 6 bits of feedback used
in~\cite{Gupt1905:Predictive}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.66\columnwidth]{images/qerror.pdf}
\caption{\label{fig:error_decay}Comparison of the $4\times 2$
  precoder quantization error vs. time for the proposed
  approach (labelled ``Cost207\_RA Scalar'') and direct Stiefel
  manifold quantization (``Cost207\_RA Manifold'') for the COST207 channel.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.71\columnwidth]{images/1024final_withtime}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 1024
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 3.41\times 10^{-6}$, corresponding to a velocity of 37.8 km/h.}
\label{fig:ber_ped}
\end{center}
\end{figure}

@@

Fig.~\ref{fig:ber_veh} shows a similar comparison for a situation with
a higher speed. The faster variation of the channel taps with time
necessitates the use of a smaller number of subcarriers (viz. 64
subcarriers), thereby permitting more accurate tracking of precoders
on a per subcarrier basis. As in the previous case, we observe that
the scalar parameter-based approach outperforms manifold based
quantization by about 1 dB for higher SNRs. Since, on a per subcarrier
basis, the channel variation is slower, the precoder tracking is
better than the previous case. Faster channel variations would impose
limits on the efficacy of the proposed method in tracking the channel
parameters.
\begin{figure}
\begin{center}
\includegraphics[width=0.71\columnwidth]{images/64final_withtime}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 64
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 1.56\times 10^{-5}$, corresponding to a velocity of 172.8 km/h.}
\label{fig:ber_veh}
\end{center}
\end{figure}

@@

\section{Conclusion}
\label{section4}
We have presented a predictive quantization based precoder
reconstruction for MIMO-OFDM systems that employs scalar quantization
based feedback of precoder parameters. By quantizing and adapting the
Givens rotation and Householder transformation parameters of the
precoder, the receiver feedback to the transmitter can be minimized
while retaining effective precoder adaptation.  Exploiting the
temporal and frequency correlations jointly enhances performance
significantly. Simulations reveal that the proposed approach provides
faster convergence and smaller errors when compared to approaches that
involve direct quantization on manifold tangent spaces. With smaller
quantization error as compared to manifold based approaches, the
feedback results in more accurate precoder tracking, along with simple
2D-interpolation over the scalar parameters to optimize the precoders.
Future work would involve tuning the adaptation rate of the scalar
parameters and further reduction of quantization redundancy to reduce
the feedback rates.

>>>>

\def\imwidth{0.65}
%package list
\IEEEoverridecommandlockouts
\let\labelindent\relax
% \usepackage{breqn}
\usepackage{enumitem}
\usepackage{fleqn}
\usepackage{cite}
\usepackage{graphicx}
\usepackage[varg]{newtxmath}
\graphicspath{ {images/} }
\usepackage{pdfpages}
\usepackage{wrapfig}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{lettrine}
\usepackage{amsmath}
% \usepackage{titlesec}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{float}
\usepackage[font={footnotesize}]{caption}
\usepackage[numbers,sort,square,compress]{natbib}
\usepackage[para]{footmisc}
\usepackage{xcolor}
\setlength{\belowcaptionskip}{-12pt}
\newcommand{\highlight}[1]{%
  \colorbox{red!50}{$\displaystyle#1$}}

@@

The parameterization approach described above is related to the channel
state information feedback present in several wireless standards,
including 802.11ac~\cite{lou2013comparison}, although adaptive
quantization and feedback are not discussed. Further
sections show that the use of adaptive approaches can greatly enhance
performance with only very minor modifications to the system.
\subsection{Differential quantization and channel tracking}
\label{quantiz}
For channels that vary slowly with time, adaptive differential
quantization is an effective method for tracking parameters over time
using very few bits (often just one bit per parameter). For example,
while tracking a slowly varying discrete random process $x_n$ at time
$n$, the one bit can be used to determine the direction in which to
move in order to track the parameter. This can be denoted by
$\beta_{n} = \mbox{sign}(x_{n} - \hat{x}_{n-1})$, where $x_n$ is the
unquantized (accurate) value and $\hat{x}_{n-1}$ is the quantized
value. Further, the step size for the move can be adapted based on the
speed of channel variation as well the accuracy needed. If $\Delta_n$
is the step size, we define our adaptive quantizer as
\begin{equation}
\hat{x}_{n} = \hat{x}_{n-1} + \beta_{n}\Delta_{n} \mbox{ where }
\label{delta_eqn}
\Delta_{n} = \begin{cases}
    M \Delta_{n-1}, & \text{if $\beta_{n} = \beta_{n-1}$}\\
    \Delta_{n-1}/M , & \text{if $\beta_{n} \neq \beta_{n-1}$}.
  \end{cases}
\end{equation}
where $M$ is the scaling factor used during the positive and negative
moves. We initialize $\Delta_1$ as $\Delta_1 = |x_{2}-\hat{x}_1|$. This is performed separately for each of the quantized
parameters.

@@

\section{Simulation and discussion}
\label{section3}
We now present simulations of the scalar feedback based joint
time-frequency predictive quantization outlined in the previous
section. In particular, we analyze the performance of the proposed
quantization and interpolation method for time-varying MIMO-OFDM
systems. The channel is simulated using the typical Jakes Model
for the COST207 channel power delay
profiles~\cite{molisch2006cost259}. The BER achieved by
the proposed predictive quantization based precoder is compared
against the ideal precoder for 100 channel evolutions in each run and
averaged over 10 channel realizations (translating to averaging over $250,000$ bits). The error properties obtained by simulating transmission over these varying channels is sufficient to obtain good average performance properties. Simulations are performed for
normalised Doppler values of $f_dT_s = 3.41\times 10^{-6}$ (corresponding
to a velocity of 37.2 km/h for $f_c = 1$ GHz) with 1024 subcarriers, and with
$f_dT_s = 1.56 \times 10^{-5}$ (velocity of 172.8 km/h) with 64
subcarriers. The sampling time period $T_s$ was $5\times10^{-8}$
s. For both the situations, we considered $N_T=4$ transmit antennas
and and $N_R=2$ receive antennas. Therefore, using the approach
outlined in~\cite{4114278}, the parameters that determine the precoder is
$N_{R}(2N_{T} - N_R) = 12$, with five $\theta_{k,l}$s and and seven
$\phi_{k,l}$s. We assume that the spacing between fed back subcarrier
indices $p$ to be $33$ and $9$ for the slow and fast channel profiles
respectively (i.e. for the slower channels with 1024 subcarriers, the
subcarriers indices fed back would be $0, 33, 66, \ldots 1023$ and for
the faster channel, with 64 subcarriers, $0, 9, 18, \ldots 63$ would
be fed back).

For the first and the second simulation time instances, since the
precoder is not available at the transmitter, we use 2 bits for
initializing each parameter for effective representation of the
precoder. Therefore, the number of bits for each subcarrier =
$10\times 2$ = 20 (we note that $\phi_0 = \phi_4 = 0$ due to the
non-uniqueness of the SVD). Quantization is performed as given in
Section~\ref{section2}. We note that this is not significant, since,
amortized over long durations, this number becomes small. Subsequent
quantization takes place in the time-domain using one bit for each
parameter. Since the bit budget is small, if the channel varies
sufficiently slowly, the $\theta_{k,l}$ and the $\phi_{k,l}$ parameters
are fed back alternately, while retaining the unsent parameters as
their previous values, as shown in
Fig.~\ref{fig:adpm-fig}.
% Here, $\bf{\Theta}$ is the collection of all
% $\theta_l$s and $\bf{\Phi}$ is the collection of all $\phi_i$s.
This brings down the bit budget to an average of 5 bits for
$\bf{\Theta}$ and 5 bits for $\bf{\Phi}$ over alternate feedback
instances, thereby resulting in an average feedback budget of 5 bits,
as opposed to the 6 bits in~\cite{Gupt1905:Predictive}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\imwidth\columnwidth]{images/qerror.pdf}
\caption{\label{fig:error_decay}Comparison of the $4\times 2$
  precoder quantization error vs. time for the proposed
  approach (labelled ``Cost207\_RA Scalar'') and direct Stiefel
  manifold quantization (``Cost207\_RA Manifold'') for the COST207 channel.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\imwidth\columnwidth]{images/1024final_withtime}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 1024
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 3.41\times 10^{-6}$, corresponding to a velocity of 37.8 km/h.}
\label{fig:ber_ped}
\end{center}
\end{figure}

@@

Fig.~\ref{fig:ber_veh} shows a similar comparison for a situation with
a higher speed. The faster variation of the channel taps with time
necessitates the use of a smaller number of subcarriers (viz. 64
subcarriers), thereby permitting more accurate tracking of precoders
on a per subcarrier basis. As in the previous case, we observe that
the scalar parameter-based approach outperforms manifold based
quantization by about 1 dB for higher SNRs. Since, on a per subcarrier
basis, the channel variation is slower, the precoder tracking is
better than the previous case. Faster channel variations would impose
limits on the efficacy of the proposed method in tracking the channel
parameters.
\begin{figure}
\begin{center}
\includegraphics[width=\imwidth\columnwidth]{images/64final_withtime}
\caption{BER vs. SNR for QPSK transmission over the $4\times 2$ 64
  subcarrier MIMO-OFDM system, with the ITU Cost207-RA channel profile
  for $f_dT_s = 1.56\times 10^{-5}$, corresponding to a velocity of 172.8 km/h.}
\label{fig:ber_veh}
\end{center}
\end{figure}

@@

\section{Conclusion}
\label{section4}
We have presented a predictive quantization based precoder
reconstruction for MIMO-OFDM systems that employs scalar quantization
based feedback of precoder parameters. Quantizing and adapting the
Givens rotation and Householder transformation parameters of precoders
minimizes receiver feedback to the transmitter while retaining
effective precoder adaptation. Exploiting the temporal and frequency
correlations jointly enhances performance significantly. Simulations
reveal that the proposed approach provides faster convergence and
smaller errors when compared to manifold based approaches. Therefore,
our feedback approach with simple 2D-interpolation over the
scalar parameters to optimize the precoders tracks precoders
effectively.  Future work would involve tuning the adaptation rate of
the scalar parameters and further reduction of quantization redundancy
to reduce the feedback rates.
